Describing the scale and composition of calls for police service: a replication and extension using open data

### 2022-02-14 paper.Rmd
time updates

\dag Netherlands Institute for the Study of Crime and Law Enforcement (NSCR), Amsterdam, Netherlands. \linebreak
\* Department of Sociology, Utrecht University, Utrecht, Netherlands \linebreak \linebreak \linebreak \linebreak

@@

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a direct comparison between different measures of 'officer time', namely, total deployment time and time on scene. These measures have been reported separately in the existing analyses, so the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g., built environment and sociodemographic backdrop). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

Recognizing that frequency counts of calls for service do not necessarily reflect the amount of police resource consumed by public demand, raw counts were augmented with the time spent by officers on each of the six classifications of demand. This information permits the calculation of a proportional breakdown of police time consumed by the different demand classifications throughout the year. The raw numbers were reported in a table. A visual representation of call types nested within each demand classification was summarized in a tree map graphic. The time committed by officers was calculated from when the call was dispatched to an officer to when the officer indicated that the event had been closed. Here, this is referred to as the _total deployed time_. The aggregate time consumed by different demand classifications was reported using a single count of officer shifts (7.5 hours) over the year.

@@

To describe the scale and composition of emergency police demand across the US more broadly, @lum2021can used CAD data from nine different police agencies. The agencies themselves were anonymized. Individual incidents were categorized into fourteen different demand classifications, namely, alarms, disorders, vice, domestic, follow-ups and service requests, mental, medical, missing persons, violence, interpersonal-other, property, suspicions, traffic, and admin, agency and non-crime. The authors reported a proportional breakdown of call counts and police time spent across the nine agencies according to the fourteen demand classifications, including the average (mean). For each agency, the authors also reported the average time spent on each demand classification in minutes, and the average across the nine agencies. No descriptive statistics on the range or spread of these figures was reported. The 'time spent' measure used was the _time on scene_. The authors considered this to be preferable to _total deployed time_ when measuring police demands, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

The data used for this study cover the city of Detroit in Michigan, US. The City of Detroit authority publish 911 emergency calls for service data through their Open Data Portal.^[Accessible via https://data.detroitmi.gov/.] In alignment with @ratcliffe2021policing, the data used here are subset for the year 2019 and exclude those calls initiated by an officer in order to capture public-initiated calls. Each individual incident has a corresponding timestamp and location coordinates to define the call in time and space.

For each incident, the police response time and time on the scene are reported. In sum, these comprise the total deployed time officers spend on the incident. Calls which had a _total deployed time_ (response time + time on scene) or _time on scene_ of zero minutes, negative values or missing were excluded. The data permit a distinction between the time spent resolving incidents with and without the response time. As noted, @lum2021can considered the _time on scene_ measure to be preferable. Here, for completeness, both measurements are reported.

@@

A series of descriptive statistics summarize the time consumed by each demand type. As an aggregate measure, we report the sum time spent on each demand classification by the number of officer shifts.^[Consistent with Ratcliffe (2021) a shift is defined as 7.5 hours. The sum total of hours for each demand classification over the year is divided by 7.5.] For comprehensiveness and transparency, we also report the mean, median, standard deviation, and range in the Appendix, next to a histogram of the underlying distribution in minutes. This sheds on light on how long a 'typical' call might take while noting a probable skew in the distribution. Such a skew would suggest that averages of time [e.g., @lum2021can] or aggregate measures of time [e.g., @ratcliffe2021policing] in isolation are insufficient and potentially mask important underlying variation, namely, that a small number of high-demand calls drag up the mean and inflate aggregate measures of time. To investigate this distribution, we plot Lorenz curves for each demand classification. Lorenz curves plot the cumulative distribution of a variable (in this case, time on scene) against the cumulative distribution of a unit (in this case, calls) [@bernasco2017more]. By way of example, we report on these figures according to arbitrary thresholds of 25% and 50% of time spent. This sheds light on the degree to which a disproportionately large amount of officer time might be consumed by just a small number of calls.

@@

## Data and methods summary

To summarize the various descriptions of police demand replicated and/or extended in this study, a checklist is outlined in Table 1. Broadly, this demonstrates that, with the exception of the final disposition, this study provides a comprehensive replication of the analyses reported by @ratcliffe2021policing and @lum2021can. Either identical descriptive statistics are reported in replication of one or both existing studies, or additional analyses are provided for enhanced insight and transparency.

```{r table1}
options(knitr.kable.NA = '')

table1_df <- read_xlsx("results/table1_methods_comparison.xlsx")

kable(table1_df, caption = "Summary of the various descriptions of demand replicated or extended in this study. Measures were reported for demand classification unless otherwise stated.",
      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
    kable_styling(latex_options = c("striped"),
                  font_size = 10, full_width = F) %>%
  column_spec(column = c(2,3,4), width = c("2cm", "2cm", "2cm") ) %>%
  row_spec(row = 14, hline_after = 14) %>%
  row_spec(row = 15, italic = T)
```

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 2. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consumed 47% of total deployed time throughout the year, although the figure increases to 51% when considering time on scene in isolation. This likely reflects the fact that crime-related calls have higher priority, and in turn, a swifter response time. The remainder of total time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%).

```{r table2}
table2a_df <- read_csv("results/table2_des_stats_total_time.csv")
table2b_df <- read_csv("results/table2_des_stats_tos.csv")

table2_df <- table2a_df %>% 
  select(-`Demand type`, -Count, -`Count (%)`) %>%  
  bind_cols(table2b_df) %>% 
  select(`Demand type`, Count, `Count (%)`,
         `Time on scene (%)`, `Total deployed time (%)`,  `Total deployed time (shifts)`) %>%  # reorder for final table.
  rename( `Total deployed \n time (shifts)` =  `Total deployed time (shifts)`)

kable(table2_df, caption = "Breakdown of frequency counts and time spent on each demand type. Total deployed time is the sum of the time on scene and response time.",
      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
    kable_styling(latex_options = c("striped"),
                  font_size = 10, full_width = F) %>% 
  column_spec(column = 1:6, width = c("2cm","2cm","2cm","2.5cm","2.5cm","2.5cm"))
```

@@

To further quantify the resources consumed, Table 3 summarizes the time spent on the scene for each demand classification in minutes. Histograms of the underlying distributions are reported in the Appendix. Table 3 demonstrates a number of things. First, there is a modest amount of variation in the typical time at scene of different types of emergency call. For instance, the mean time spent on the scene of traffic incidents (55 minutes) and crime (51 minutes) is considerably more that quality of life issues (31 minutes) or proactive activity (28 minutes). These differences will be a result of a combination of factors, including the complexity and severity of incidents and the number of officers required to attend the scene. Second, there is considerable variation in the time consumed by incidents _within_ each demand type. While many calls are resolved in minutes, a handful consume hours of cumulative officer time, resulting in a highly skewed distribution. This highlights how mean or aggregate summary statistics of police time as a measure of demand can mask (important) underlying variation.

```{r table3}
table3_df <- read_csv("results/table3_des_stats_tos.csv")

kable(table3_df, caption = "Descriptive statistics of minutes spent on scene for each demand type classification.",
      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
    kable_styling(latex_options = c("striped"),
                  font_size = 10, full_width = F) %>% 
  column_spec(column = 1:6, width = c("2.5cm","1cm","1cm","1cm","1.2cm", "1cm"))
```

@@

Figure 2 visualizes the disproportionate share of time spent on a small number of calls explicitly using Lorenz curves for each demand classification. These can be compared against the reference line of perfect equality. Here, the line of perfect equality represents a situation in which officer time is distributed evenly across all calls. In reality, we can see that this is not the case: the Lorenz curves for each demand classification deviate considerably from perfect equality. The curves for each classification are almost indistinguishable, so we focus on the curve for all calls (in blue). This shows that 50% of the total time officers spent on the scene during 2019 was consumed by just 15% of calls, and 25% of time on scene was consumed by just 5% of all calls. In other words, a small amount of emergency calls for service account for a disproportionately large amount of police resources out on the street. This finding is consistent (and comparable) across all six classifications of demand.

@@

In this study, calls were categorized according the six classes used by @ratcliffe2021policing. The overall composition of police demand in terms of call volume and proportional time spent are remarkably similar. A considerable volume of calls for service, and in turn, a considerable proportion of deployed police time, relate to incident types which do not fall under the traditional 'crime-fighting' role of the police. Instead, the police fulfill a variety of other social service functions, including the resolution of public (mental) health incidents and resolving community issues. That said, descriptive statistics which drill down into the underlying distribution of officer time consumed by each demand type provide additional insight which would otherwise be overlooked. First, it is clear that there is considerable variation in the amount of time consumed both between and within demand classifications. On a methodological point, this highlights how average (e.g., mean times) or aggregate (e.g., sum times) might mask underlying variation. Second, and relatedly, these extensions demonstrate that a disproportionately large amount of police time is spent on a small number of emergency calls. 50% of total time on scene in consumed by just 15% of calls. High-demand calls will likely be complex and/or life-threatening incidents which require numerous officers to be present at the scene, or for extended periods of time. A targeted examination of these calls might shed light on the extent to which such high-intensity demands could be reduced or better resolved through or in concert with other public services. Targeted interventions on just a handful of calls could drastically reduce the public demand for police services.

>>>>

\begin{flushleft}

\dag Netherlands Institute for the Study of Crime and Law Enforcement (NSCR), Amsterdam, Netherlands. \linebreak
*Department of Sociology, Utrecht University, Utrecht, Netherlands \linebreak \linebreak \linebreak \linebreak

\end{flushleft}

@@

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a comparison between measures of 'officer time' with and without the response time. While concerns have been raised about the inclusion of the response time in measures of officer time, largely due to low-priority calls not necessitating a rapid response, the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g., built environment). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

Recognizing that frequency counts of calls for service do not necessarily reflect the amount of police resource consumed by public demand, raw counts were augmented with the time spent by officers on each of the six classifications of demand. This information permits the calculation of a proportional breakdown of police time consumed by the different demand classifications throughout the year. The raw numbers were reported in a table. A visual representation of call types nested within each demand classification was summarized in a tree map graphic. The time committed by officers was calculated from when the call was dispatched to an officer to when the officer indicated that the event had been closed. Ratcliffe accounted for the number of officers attending each incident when calculating the amount of time consumed. The aggregate time consumed by different demand classifications was reported using a single count of officer shifts over the year.

@@

To describe the scale and composition of emergency police demand across the US more broadly, @lum2021can used CAD data from nine different police agencies. The agencies themselves were anonymized. Individual incidents were categorized into fourteen different demand classifications, namely, alarms, disorders, vice, domestic, follow-ups and service requests, mental, medical, missing persons, violence, interpersonal-other, property, suspicions, traffic, and admin, agency and non-crime. The authors reported a proportional breakdown of call counts and police time spent across the nine agencies according to the fourteen demand classifications, including the average (mean). For each agency, the authors also reported the average time spent on each demand classification in minutes, and the average across the nine agencies. No descriptive statistics on the range or spread of these figures was reported. The 'time spent' measure used was the _time on scene_ (i.e., excluding response time). The authors considered this to be preferable, since low priority calls may have a non-urgent (slow) response due to the nature of the call, and officers may be reassigned to other duties while on an assignment. Lum and colleagues calculated _time on scene_ as the time from the first officer's arrival to when the call was reported complete. In this way, the calculation did not account for multi-officer responses (e.g., back-up). Instead, the authors calculated _time on scene_ as the time from the first officer's arrival to when the call was reported complete. This was justified for a number of reasons, among them, avoiding overinflation of time due to officers 'flocking' to non-serious incidents, and limitations with the CAD data. A proportional breakdown of the disposition (assist only, report written or arrest citation) was provided for each of the fourteen call types. No spatial or temporal findings were reported.

@@

The data used for this study cover the city of Detroit in Michigan, US. The City of Detroit authority publish 911 emergency calls for service data through their Open Data Portal.^[Accessible via https://data.detroitmi.gov/.] In alignment with the two existing studies, the data used here exclude those calls initiated by an officer in order to capture public-initiated calls. Each individual incident has a corresponding timestamp and location coordinates to define the call in time and space. Data was subset for the year 2019.

For each incident, the police response time and time on the scene are reported. The former measures the response time of the first officer on-scene. The latter mimics the measure used by @lum2021can, namely, the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. Here, we refer to the sum of the response time and time on scene as the _total deployed time_. Calls which had a _total deployed time_ or _time on scene_ of zero minutes, negative values or missing were excluded. The data permit a distinction between the time spent with and without the response time. As noted, @lum2021can considered the _time on scene_ measure to be preferable. Here, for completeness, we report times with and without the response time.

@@

A series of descriptive statistics summarize the time on scene consumed by each demand type. For comprehensiveness and transparency, we also report the mean, median, standard deviation, and range in the Appendix, next to a histogram of the underlying distribution in minutes. This sheds on light on how long a 'typical' call might take while noting a probable skew in the distribution. Such a skew would suggest that averages of time [e.g., @lum2021can] or aggregate measures of time [e.g., @ratcliffe2021policing] in isolation are insufficient and potentially mask important underlying variation, namely, that a small number of high-demand calls drag up the mean and inflate aggregate measures of time. To investigate this distribution, we plot Lorenz curves for each demand classification. Lorenz curves plot the cumulative distribution of a variable (in this case, time on scene) against the cumulative distribution of a unit (in this case, calls) [@bernasco2017more]. By way of example, we report on these figures according to arbitrary thresholds of 25% and 50% of time spent. This sheds light on the degree to which a disproportionately large amount of officer time might be consumed by just a small number of calls.

<!-- As an aggregate measure, we report the sum time spent on each demand classification by the number of officer shifts.^[Consistent with Ratcliffe (2021) a shift is defined as 7.5 hours. The sum total of hours for each demand classification over the year is divided by 7.5.]  -->

@@

<!-- ## Data and methods summary -->

<!-- To summarize the various descriptions of police demand replicated and/or extended in this study, a checklist is outlined in Table 1. Broadly, this demonstrates that this study provides a comprehensive replication of the analyses reported by @ratcliffe2021policing and @lum2021can. Either identical descriptive statistics are reported in replication of one or both existing studies, or additional analyses are provided for enhanced insight and transparency. Here, we note two exceptions: we do not report an aggregate summary (total shifts) because our time measure does not  -->

<!-- \newpage -->

<!-- ```{r table1} -->
<!-- options(knitr.kable.NA = '') -->

<!-- table1_df <- read_xlsx("results/table1_methods_comparison.xlsx") -->

<!-- kable(table1_df, caption = "Summary of the various descriptions of demand replicated or extended in this study. Measures were reported for demand classification unless otherwise stated.", -->
<!--       booktabs = T, longtable = T, linesep = "", format = "latex") %>% -->
<!--     kable_styling(latex_options = c("striped"), -->
<!--                   font_size = 10, full_width = F) %>% -->
<!--   column_spec(column = c(2,3,4), width = c("2cm", "2cm", "2cm") ) %>% -->
<!--   row_spec(row = 14, hline_after = 14) %>% -->
<!--   row_spec(row = 15, italic = T) -->
<!-- ``` -->

@@

The frequency counts and proportional breakdowns for each demand type classification are reported in Table 1. This demonstrates that in Detroit, a considerable proportion of citizen calls for service, and in turn, time spent by officers, is committed to a diverse array of (often non-criminal) issues. Just 44% of emergency calls for service related to crime incidents. These crime incidents consumed 47% of total deployed time throughout the year, although the figure increases to 51% when considering time on scene in isolation. This likely reflects the fact that crime-related calls have higher priority, and in turn, a swifter response time.^[Crime calls for service had the highest proportion of Priority 1 (highest priority) responses at 52%, followed by traffic (24%), quality of life (17%), community (16%), health (14%), and proactive (3%).] The remainder of total time on scene is spent dealing with traffic duties (16%), quality of life incidents (15%), public and mental health (7%), community issues (5%), and proactive policing (4%).

```{r table1}
table1a_df <- read_csv("results/table1_des_stats_total_time.csv")
table1b_df <- read_csv("results/table1_des_stats_tos.csv")

table1_df <- table1a_df %>% 
  select(-`Demand type`, -Count, -`Count (%)`) %>%  
  bind_cols(table1b_df) %>% 
  select(`Demand type`, Count, `Count (%)`, # ,  `Total deployed time (shifts)`
         `Time on scene (%)`, `Total deployed time (%)`) #%>%  reorder for final table.
  # rename( `Total deployed \n time (shifts)` =  `Total deployed time (shifts)`)

kable(table1_df, caption = "Breakdown of frequency counts and time spent on each demand type. Total deployed time is the sum of the time on scene and response time.",
      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
    kable_styling(latex_options = c("striped"),
                  font_size = 10, full_width = F) %>% 
  column_spec(column = 1:6, width = c("2cm","2cm","2cm","2.5cm","2.5cm","2.5cm"))
```

@@

To further quantify the resources consumed, Table 2 summarizes the time spent on the scene for each demand classification in minutes. Histograms of the underlying distributions are reported in the Appendix. Table 2 demonstrates a number of things. First, there is a modest amount of variation in the typical time at scene of different types of emergency call. For instance, the mean time spent on the scene of traffic incidents (55 minutes) and crime (51 minutes) is considerably more that quality of life issues (31 minutes) or proactive activity (28 minutes). These differences will be a result of a combination of factors, including the complexity and severity of incidents. Second, there is considerable variation in the time consumed by incidents _within_ each demand type. While many calls are resolved in minutes, a handful consume hours of officer time on the scene, resulting in a highly skewed distribution. This highlights how mean or aggregate summary statistics of police time as a measure of demand can mask (important) underlying variation.

```{r table2}
table2_df <- read_csv("results/table2_des_stats_tos.csv")

kable(table2_df, caption = "Descriptive statistics of minutes spent on scene for each demand type classification.",
      booktabs = T, longtable = T, linesep = "", format = "latex") %>%
    kable_styling(latex_options = c("striped"),
                  font_size = 10, full_width = F) %>% 
  column_spec(column = 1:6, width = c("2.5cm","1cm","1cm","1cm","1.2cm", "1cm"))
```

@@

Figure 2 visualizes the disproportionate share of time spent on a small number of calls explicitly using Lorenz curves for each demand classification. These can be compared against the reference line of perfect equality. Here, the line of perfect equality represents a situation in which officer time is distributed evenly across all calls. In reality, we can see that this is not the case: the Lorenz curves for each demand classification deviate considerably from perfect equality. The curves for each classification are almost indistinguishable, so we focus on the curve for all calls (in blue). This shows that 50% of the time officers spent on the scene during 2019 was consumed by just 15% of calls, and 25% of time on scene was consumed by just 5% of all calls. In other words, a small amount of emergency calls for service account for a disproportionately large amount of police resources out on the street. This finding is consistent (and comparable) across all six classifications of demand.

@@

In this study, calls were categorized according the six classes used by @ratcliffe2021policing. The overall composition of police demand in terms of call volume and proportional time spent are remarkably similar. A considerable volume of calls for service, and in turn, a considerable proportion of deployed police time, relate to incident types which do not fall under the traditional 'crime-fighting' role of the police. Instead, the police fulfill a variety of other social service functions, including the resolution of public (mental) health incidents and resolving community issues. That said, descriptive statistics which drill down into the underlying distribution of officer time consumed by each demand type provide additional insight which would otherwise be overlooked. First, it is clear that there is considerable variation in the amount of time consumed both between and within demand classifications. On a methodological point, this highlights how average (e.g., mean times) or aggregate (e.g., sum times) might mask underlying variation. Second, and relatedly, these extensions demonstrate that a disproportionately large amount of police time is spent on a small number of emergency calls. 50% of total time on scene in consumed by just 15% of calls. High-demand calls will likely be complex and/or life-threatening incidents or calls which require officers to secure a scene for extended periods of time. A targeted examination of these calls might shed light on the extent to which such high-intensity demands could be reduced or better resolved through or in concert with other public services.

On this point, this study raises useful discussion points around the different measures of officer time. First, it is clear that findings on the scale and composition of police demand are sensitive, albeit only slightly, to whether measures of officer time include the response time. The degree to which the two estimates differ varies by demand classification, likely as a result of the different call priorities, and in turn, the rapidness of the response times. That said, this claim would need verifying through detailed examination. On this point, second, we would reiterate that the 'time on scene' measure reported here, in alignment with @lum2021can, refers to the amount of time which elapses between when the first officer arrives and when the last unit clears the scene. This differs to that of @ratcliffe2021policing which accounted the number of officers attending the scene. @lum2021can note that including back-up officers in the time calculation might introduces complications. For instance, non-serious incidents might receive back-up unnecessarily, particularly during periods of low-demand, which would inflate the time consumed and overstate the resources required to resolve such an incident. Although the overall composition of police demand is comparable between Philadelphia [@ratcliffe2021policing] and Detroit, the true impact of these measures remains unclear. We propose that future research investigates these distinctions. This would require CAD (or equivalent) data, timestamped at each stage the response (call-handler dispatch, arrival on scene, scene departure), for individual police units nested within calls. This level of granularity would permit the calculation and comparison of multiple measures of officer time, as well as an investigation into the extent to which back-ups and/or 'flocking' might influence such measures.
